Data Migration under a Changing Schema in Ampersand

### 2023-11-13 2022Migration/articleMigrationFACS.tex
two new figures

\section{Introduction}
\label{sct:Introduction}
   In practice, information systems%
   \footnote{In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.}
   may live for many years.
   After they are built, they need to be updated regularly to keep up with changing requirements in a dynamically evolving environment.
   Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
   Obviously, these deployments are mostly updates of existing systems.
   Schema changes cannot always be avoided when updating software, so a {\em schema changing data migration} (SCDM) will be neccessary from time to time.
   For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
   Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
   The risk and effort of such data migrations explains why development teams try to avoid schema changes.

@@

   In practice, data migrations typically follow Extract-Transform-Load (ETL) patterns~\cite{Theodorou2017},
   for which many tools are available.
   However, invariants that change require writing code in the transform phase,
   for which ETL tools typically provide little support.
   This yields extra work for software engineers, and it increases the risk of software errors.
   The ubiquitous pressure to decrease deploy times calls for further automation of SCDMs.
   With the release frequencies increasing and deploy times decreasing, any downtime for the sake of migration becomes less and less acceptable.
   That is why our research aims for zero downtime.

   Another practical problem is that of data quality.
   Migrations typically suffer from a backlog of deteriorated data, incurring work to clean it up.
   Some of that work must be done before the migration; some can wait till after the migration.
   We can capture part of the data quality problem as a requirement to satisfy semantic constraints.
   E.g. the constraint that the combination of street name, house number, postal code, and city occurs in
   a registration of valid addresses can be checked automatically.
   In a formalism like Ampersand, which allows us to express such constraints, we can add constraints for data quality to the schema.
   This allows us to signal the data pollution at runtime.
   Some forms of data pollution cannot be detected in this way, however.
   An example is when a person has deliberately specified a false name without violating any constraint in the system.

@@

\section{Analysis}
\label{sct:Analysis}
   This section analyzes information systems qualitatively, to prepare for a formal treatment in section~\ref{sct:Definitions}.
   The current section yields a procedure for migrating data from one system to another.
\subsection{Information Systems}
   The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
   Multiple users, working from different locations and on different moments, constitute what we will loosely call ``the business''.
   The data in the system constitutes the collective memory of the business,
   which relies on the semantics of the data to draw the right conclusions and carry out their tasks.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.45]{figures/datamigration-Pre-migration.png}
   \end{center}
\caption{Anatomy of an information system}
\label{fig:pre-migration}
\end{figure}

   Actors (both users and computers) are changing the data in a system continually.
   The state of the system is represented by a data set, typically represented in some form of persistent store such as a database.
   Events that the system detects may cause the state to change.
   To keep our theory technology independent, we assume data sets to contain triples.
   This makes our theory valid for every kind of database that triples can represent,
   including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.

@@

\subsection{Data Migrations}
   Data migration occurs when a desired system replaces an existing one,
   while preserving the meaning of the present data as much as possible~\cite{Spivak2012}.
   %Just copying the set of data from the existing system to the desired system is obviously wrong if the schemas of both systems differ.
   %
   In practice, data migrations typically deploy the existing system and the desired system side-by-side,
   while transferring data in a controlled fashion, as shown in Figure~\ref{fig:migration phase}.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.35]{figures/datamigration-Migration-phase.png}
   \end{center}
\caption{Migration phase}
\label{fig:migration phase}
\end{figure}
   To automate the migration as much as possible and to achieve zero downtime,
   we must deploy a third system: the migration system.
   This system has its own schema.
   It comprises the schemas of both the existing system and the desired system,
   so the migration system can copy the data from the existing to the desired system.
   It uses transactions to copy the data and to resolve some forms of data pollution.
   Not all migration work, however, can be automated.
   Data pollution, new business rules, or known issues in the existing system
   may occasionally require tailoring the migration script to specific needs
   that require action of users in production.
   For that purpose, the migration engineer specifies business constraints in the migration system.
   In Ampersand, a developer can use business constraints to implement such needs.

>>>>

\section{Introduction}
\label{sct:Introduction}
   In practice, information systems%
   \footnote{In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.}
   may live for many years.
   After they are built, they need to be updated regularly to keep up with changing requirements in a dynamically evolving environment.
   Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
   Obviously, these deployments are mostly updates of existing systems.
   Schema changes cannot always be avoided when updating software, so a {\em schema changing data migration} (SCDM) will be neccessary from time to time.
   For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
   Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
   In practice, data migrations typically follow Extract-Transform-Load (ETL) patterns~\cite{Theodorou2017},
   for which many tools are available.
   However, ETL tools typically provide little support for invariants that change, forcing development teams to write code.
   The risk and effort of such data migrations explains why these teams try to avoid schema changes.
   Our research aims at (partly) automating SCDMs to make them less risky and less costly,
   so development teams can increase the frequency of releases and do schema changes with zero downtime.

@@

   Another practical problem is that of data quality.
   Migrations typically suffer from a backlog of deteriorated data, incurring work to clean it up.
   Some of that work must be done before the migration; some can wait till after the migration.
   We can capture the automatable part of the data quality problem by regarding it as a requirement to satisfy semantic constraints.
   E.g. the constraint that the combination of street name, house number, postal code, and city occurs in
   a registration of valid addresses can be checked automatically.
   In a formalism like Ampersand, which allows us to express such constraints, we can add constraints for data quality to the schema.
   This allows us to signal the data pollution at runtime.
   Some forms of data pollution are not automatable, however.
   An example is when a person has deliberately specified a false name without violating any constraint in the system.

@@

\section{Analysis}
\label{sct:Analysis}
   This section analyzes information systems qualitatively, to prepare for a formal treatment in section~\ref{sct:Definitions}.
   The current section yields a procedure for migrating data from one system to another.
\subsection{Information Systems}
   The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
   Multiple users, working from different locations and on different moments, constitute what we will loosely call ``the business''.
   The data in the system constitutes the collective memory of the business,
   which relies on the semantics of the data to draw the right conclusions and carry out their tasks.
   \begin{figure}[bht]
      \begin{center}
        \includegraphics[scale=0.8]{figures/existing system.pdf}
      \end{center}
   \caption{Anatomy of an information system}
   \label{fig:pre-migration}
   \end{figure}

   Figure~\ref{fig:pre-migration} depicts the situation before migration.
   An existing application service ingests traffic through an ingress and persists data in a data set, which is typically a database.
   This research assumes that the structure and business semantics are represented in a schema, from which the system is generated.
   Actors (both users and computers) are changing the data in a system continually.
   The state of the system is represented by a data set, typically represented in some form of persistent store such as a database.
   Events that the system detects may cause the state to change.
   To keep our theory technology independent, we assume data sets to contain triples.
   This makes our theory valid for every kind of database that triples can represent,
   including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.

@@

\subsection{Data Migrations}
   Data migration occurs when a desired system replaces an existing one,
   while preserving the meaning of the present data as much as possible~\cite{Spivak2012}.
   %Just copying the set of data from the existing system to the desired system is obviously wrong if the schemas of both systems differ.
   %
   In practice, data migrations typically deploy the existing system and the desired system side-by-side,
   while transferring data in a controlled fashion, as shown in Figure~\ref{fig:migration phase}.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.8]{figures/migration system deployed.pdf}
   \end{center}
\caption{Migration phase}
\label{fig:migration phase}
\end{figure}
   To automate the migration as much as possible and to achieve zero downtime,
   we must deploy a third system: the migration system.
   This system has its own schema.
   It comprises the schemas of both the existing system and the desired system,
   so the migration system can copy the data from the existing to the desired system.
   It uses transactions to copy the data and to resolve some forms of data pollution.
   Not all migration work, however, can be automated.
   Data pollution, new business rules, or known issues in the existing system
   may occasionally require tailoring the migration script to specific needs
   that require action of users in production.
   For that purpose, the migration engineer specifies business constraints in the migration system.
   In Ampersand, a developer can use business constraints to implement such needs.
